---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Download and unzip the data
```{r}
# unzip the activity.zip file if activity.csv does not exist
if (!file.exists("activity.csv")) {
  unzip("activity.zip")
}
```

Read the data into a data frame and change the date format into Date
```{r}
# read activity.csv
activity.file <- read.csv("activity.csv", stringsAsFactors = FALSE, na.strings ="NA")

# convert the date (column 2) from chr to Date format
activity.file[, 2] <- as.Date(activity.file[, 2]) ## format column 2 (date) as Date
```



## What is mean total number of steps taken per day?

Remove missing values and calculate the number of steps per day.
Plot the total number of steps taken per day
```{r}
activity <- subset(na.omit(activity.file)) ## remove NA
steps.per.day <- aggregate(steps ~ date, activity, sum) ## calculate the number of steps per day

library(ggplot2)
h <- ggplot(steps.per.day, aes(x=date, y=steps)) + 
      geom_bar(stat = "identity", colour="green") + 
      ggtitle("Total number of steps taken per day") + 
      xlab("Date") + 
      ylab("Number of Steps")
print(h)
```

Compute the mean and median

```{r}
mean.steps <- mean(steps.per.day$steps) ## compute the mean of steps per day
mean.steps
median.steps <- median(steps.per.day$steps)## compute the median of steps per day
median.steps
```

### The mean is `r mean.steps` and the median is `r median.steps`.


## What is the average daily activity pattern?
 Calculate the mean steps in the 5-minute intervals and then plot the time series
```{r}
steps.interval <- aggregate(steps ~ interval, activity, mean) ## calculate the mean steps in the 5-min intervals

j <- ggplot(steps.interval, aes(interval)) + 
     geom_line(aes(y=steps)) +
     ggtitle("Average number of steps taken in the 5-minute intervals") +
     xlab("Intervals") + 
     ylab("Average number of Steps")
print(j)
```


Find which interval has the maximum number of steps

```{r}
the.interval <- steps.interval$interval[which.max(steps.interval$steps)] # determine which interval has the maximun number of steps
the.interval
```
### The interval that contains the maximum number of steps is `r the.interval`.


## Imputing missing values
Find the total number of missing values.
```{r}
total.NA <- sum(is.na(activity.file$steps))
total.NA
```
### The total number of missing values in the data aset is `r total.NA`

Impute missing values with the mean number of steps per day
```{r}
library(Hmisc)
impute.data <- activity.file
impute.data$steps <- impute(activity.file$steps, fun=mean)
new.steps.per.day <- aggregate(steps ~ date, impute.data, sum) ## sum of steps per day for impute data
```
Plot the total number of steps taken per day after missing values were imputed.

```{r}
k <- ggplot(new.steps.per.day, aes(x=date, y=steps)) + 
        geom_bar(stat = "identity", colour="blue") + 
        ggtitle("Total number of steps taken per day after imputing NAs") + 
        xlab("Date") + 
        ylab("Number of Steps")
print(k)
```

Compute the mean and median for the newly imputed data
```{r}
new.mean.steps <- mean(new.steps.per.day$steps) ## compute the mean for steps per day for impute data
new.mean.steps
new.median.steps <- median(new.steps.per.day$steps) ## compute the median of steps per day for impute data
new.median.steps
```

### The mean is `r new.mean.steps` and the median is `r new.median.steps`.


## Are there differences in activity patterns between weekdays and weekends?
Use the weekdays function to determine the type of day as either weekend or weekday
```{r}
library(dplyr)
impute.data <- mutate(impute.data, weektype = ifelse(weekdays(impute.data$date) == "Saturday" | 
                weekdays(impute.data$date) == "Sunday", "weekend", "weekday"))
```

Change the new weektype column to factor and look at the structure of the data

```{r}
impute.data$weektype <- as.factor(impute.data$weektype)
str(impute.data)
```

Change the steps type to numeric and plot the graph into panel based on weektype

```{r}
impute.data$steps <- as.numeric(impute.data$steps)

g <- ggplot(impute.data, aes(interval, steps, color = weektype)) +
        geom_line() + facet_wrap(~ weektype, ncol=1)
print(g)
```

There is a slight difference in activity pattern between weekdays and weekends.  More steps were taken during the morning of weekdays compared to weekends.
